Supporting Information for: A Standardized Effect Size for Evaluating and Comparing the Strength of Phylogenetic Signal

### 2020-08-23 Manuscript/SupportingInfo.tex
vsn 3

@@

Results from simulations on pectinate phylogenies largely mirrored those
found on pure-birth trees. For \(\hat{\lambda}\), the mean value
increased with increasing input signal, but was negatively biased, and
was less than the input value across most of its range (Fig. S1: black
line). Additionally, the precision of \(\hat{\lambda}\) varied with
differing input levels, with the greatest variation found at
intermediate values of \(\lambda\) (Fig. S1 red line). Finally, the
distribution of \(\hat{\lambda}\) was not normal, and became more skewed
at more extreme values of \(\lambda\) (Fig. S1 blue line). \hfill\break

For \(\hat\kappa\), mean values increased with increasing phylogenetic
signal, though as was found with pure-birth trees, the increase was
nonlinear (Fig. S2 black line). Likewise, variation increased with
increasing phylogenetic signal (Fig. S2 red line), though the
distribution of \(\hat\kappa\) was more normally distributed throughout
its range, and across different tree sizes, as compared with
\(\hat{\lambda}\), thought there was some slight skew for high input
levels of phylogenetic signal on large phylogenies (Fig. S2 blue line).
\hfill\break

With respect to effect sizes, both \(Z_{\lambda}\) and \(Z_{\kappa}\)
increased with increasing input phylogenetic signal, but
\(\hat{\lambda}\) was strongly affected by tree size, whereas
\(Z_{\kappa}\) was more consistent (Fig. S3). Also, \(Z_{\kappa}\)
increased more linearly with increasing levels of phylogenetic signal,
and its standard deviation across input signal was more even across tree
sizes, implying more consistent precision.

\textbf{Figure S1}. Response of Pagel's \(\lambda\) to increasing
strength of Brownian motion on pectinate trees. Gray line signifies the
1:1 line where the input value matches the estimate.

@@

Results from simulations on balanced phylogenies also mirrored those
found on pure-birth trees. For \(\hat{\lambda}\), the mean value
increased with increasing input signal, but was negatively biased, and
was less than the input value across most of its range (Fig. S4: black
line). Additionally, the precision of \(\hat{\lambda}\) varied with
differing input levels, with the greatest variation found at
intermediate values of \(\lambda\) (Fig. S4 red line). Finally, the
distribution of \(\hat{\lambda}\) was not normal, and became more skewed
at more extreme values of \(\lambda\) (Fig. S4 blue line). \hfill\break

For \(\hat\kappa\), mean values increased with increasing phylogenetic
signal, though as was found with pure-birth trees, the increase was
nonlinear (Fig. S5 black line). Likewise, variation increased with
increasing phylogenetic signal (Fig. S5 red line), though the
distribution of \(\hat\kappa\) was more normally distributed throughout
its range, and across different tree sizes, as compared with
\(\hat{\lambda}\), thought there was some slight skew for high input
levels of phylogenetic signal on large phylogenies (Fig. S5 blue line).
\hfill\break

With respect to effect sizes, both \(Z_{\lambda}\) and \(Z_{\kappa}\)
increased with increasing input phylogenetic signal, but
\(\hat{\lambda}\) was strongly affected by tree size (fig.~S6). Also,
\(Z_{\kappa}\) increased more linearly with increasing levels of
phylogenetic signal, though there was some affect of tree size with
balanced phylogenies. But as before, its standard deviation across input
signal was more even across tree sizes, implying more consistent
precision.

\textbf{Figure S4}. Response of Pagel's \(\lambda\) to increasing
strength of Brownian motion on balanced trees. Gray line signifies the
1:1 line where the input value matches the estimate.

@@

We also investigaged the effect of unresolved phylogenies on estimates
of phylogenetic signal by adjusting pure-birth trees to have 50\%
collapsed nodes, following the procedures of (1). Results from
simulations on phylogenies containing polytomies mirrored those found on
fully resolved trees. For \(\hat{\lambda}\), the mean value increased
with increasing input signal, but was negatively biased, and was less
than the input value across most of its range (Fig. S7: black line).
Additionally, the precision of \(\hat{\lambda}\) varied with differing
input levels, with the greatest variation found at intermediate values
of \(\lambda\) (Fig. S7 red line). Finally, the distribution of
\(\hat{\lambda}\) was not normal, and became more skewed at more extreme
values of \(\lambda\) (Fig. S7 blue line). \hfill\break

For \(\hat\kappa\), mean values increased with increasing phylogenetic
signal, though as was found with pure-birth trees, the increase was
nonlinear (Fig. S8 black line). Likewise, variation increased with
increasing phylogenetic signal (Fig. S8 red line), though the
distribution of \(\hat\kappa\) was more normally distributed throughout
its range, and across different tree sizes, as compared with
\(\hat{\lambda}\) (Fig. S8 blue line). \hfill\break

With respect to effect sizes, both \(Z_{\lambda}\) and \(Z_{\kappa}\)
increased with increasing input phylogenetic signal, but
\(\hat{\lambda}\) was strongly affected by tree size, whereas
\(Z_{\kappa}\) was more consistent (Fig. S9). Also, \(Z_{\kappa}\)
increased more linearly with increasing levels of phylogenetic signal,
and its standard deviation across input signal was more even across tree
sizes, implying more consistent precision. Thus, polytomies do not exert
an appreciable effect on \(Z_{\kappa}\).

\textbf{Figure S7}. Response of Pagel's \(\lambda\) to increasing
strength of Brownian motion on pure-birth trees containing polytomies.
Gray line signifies the 1:1 line where the input value matches the
estimate.

@@

We also investigated the effect of incorporating \(\lambda\) in PGLS
analyses (phylogenetic regression and ANOVA). Here, patterns for
\(\hat\lambda\), \(\hat\kappa\), \(Z_\lambda\) and \(Z_\kappa\) were
virtually identical to those found previously. In essence,
\(\hat\lambda\) was not suitable as a test statistic representing
phylogenetic signal due to all properties previously shown (Figs. S10 \&
S11), \(\hat\kappa\) was more appropriate, but associated non-linearly
with signal strength (Figs. S12 \& 13), while \(Z_\kappa\) was found to
be a superior effect size as compared with \(Z_\lambda\) (Figs S14 \&
S15). \hfill\break

With respect to model performance, input parameters (\(\beta\)) were
well estimated in phylogenetic regression and ANOVA when \(\lambda\) was
incorporated (Figs S16 \& S17), implying that \(\lambda\) may be used as
a tuning parameter for these models. Additionally, type I error and
statistical power were unaffected by the inclusion of \(\lambda\) (Figs
S18 \& S19). This latter result confirms earlier findings of (2) for
phylogenetic regression, and extends them to the case of phylogenetic
ANOVA.

\textbf{Figure S10}. Response of Pagel's \(\lambda\) to increasing
strength of Brownian motion in phylogenetic regression (incorporating
\(\lambda\)). Gray line signifies the 1:1 line where the input value
matches the estimate.

@@

\textbf{Figure S11}. Response of Pagel's \(\lambda\) to increasing
strength of Brownian motion in phylogenetic ANOVA (incorporating
\(\lambda\)). Gray line signifies the 1:1 line where the input value
matches the estimate.

@@

\textbf{Figure S12}. Response of Blomberg's \(\kappa\) to increasing
strength of Brownian motion in phylogenetic regression (incorporating
\(\lambda\)). Gray line signifies the 1:1 line where the input value
matches the estimate.

@@

\textbf{Figure S13}. Response of Blomberg's \(\kappa\) to increasing
strength of Brownian motion in phylogenetic ANOVA (incorporating
\(\lambda\)). Gray line signifies the 1:1 line where the input value
matches the estimate.

@@

\textbf{Figure S14}. Response of effect sizes \(Z_{\lambda}\) and
\(Z_{\kappa}\) to increasing strength of Brownian motion in phylogenetic
regression (incorporating \(\lambda\)).

@@

\textbf{Figure S15}. Response of effect sizes \(Z_{\lambda}\) and
\(Z_{\kappa}\) to increasing strength of Brownian motion in phylogenetic
ANOVA (incorporating \(\lambda\)).

@@

\textbf{Figure S16}. Parameter estimates from phylogenetic regression
(incorporating \(\lambda\)). Gray line is the 1:1 line and red line is
the estimate.

@@

\textbf{Figure S17}. Parameter estimates from phylogenetic ANOVA
(incorporating \(\lambda\)). Gray line is the 1:1 line and red line is
the estimate.

@@

\leavevmode\hypertarget{ref-Revell2010}{}%
2. Revell LJ (2010) Phylogenetic signal and linear regression on species
data. \emph{Methods in Ecology and Evolution} 1:319--329.

>>>>

\hypertarget{additional-results-on-pure-birth-phylogenies}{%
\subsection{Additional Results on Pure-Birth
Phylogenies}\label{additional-results-on-pure-birth-phylogenies}}

Results from the simulations on pure-birth trees revealed that both
\(Z_{\lambda}\) and \(Z_K\) increased with increasing input phylogenetic
signal, but \(\hat{\lambda}\) was strongly affected by tree size,
whereas \(Z_K\) was m1re consistent (Fig. S4). Additionally, \(Z_K\)
increased more linearly with increasing levels of phylogenetic signal,
and its standard deviation across input signal was more even across tree
sizes, implying more consistent precision. This latter observation is
seen in the upper four panels of Fig. S1, where the dispersion of
simulation results (gray dots) is far more even at each input level of
phylogenetic signal for \(Z_K\) than it is for \(Z_{\lambda}\).

\textbf{Figure S1}. Response of effect sizes \(Z_{\lambda}\) and \(Z_K\)
to increasing strength of Brownian motion.

@@

Results from simulations on pectinate phylogenies largely mirrored those
found on pure-birth trees. For \(\hat{\lambda}\), the mean value
increased with increasing input signal, but was negatively biased, and
was less than the input value across most of its range (Fig. S2: black
line). Additionally, the precision of \(\hat{\lambda}\) varied with
differing input levels, with the greatest variation found at
intermediate values of \(\lambda\) (Fig. S2 red line). Finally, the
distribution of \(\hat{\lambda}\) was not normal, and became more skewed
at more extreme values of \(\lambda\) (Fig. S2 blue line). \hfill\break

For \emph{K}, mean values increased with increasing phylogenetic signal,
though as was found with pure-birth trees, the increase was nonlinear
(Fig. S3 black line). Likewise, variation increased with increasing
phylogenetic signal (Fig. S3 red line), though the distribution of
\emph{K} was more normally distributed throughout its range, and across
different tree sizes, as compared with \(\hat{\lambda}\), thought there
was some slight skew for high input levels of phylogenetic signal on
large phylogenies (Fig. S3 blue line). \hfill\break

With respect to effect sizes, both \(Z_{\lambda}\) and \(Z_K\) increased
with increasing input phylogenetic signal, but \(\hat{\lambda}\) was
strongly affected by tree size, whereas \(Z_K\) was more consistent
(Fig. S4). Also, \(Z_K\) increased more linearly with increasing levels
of phylogenetic signal, and its standard deviation across input signal
was more even across tree sizes, implying more consistent precision.

@@

Results from simulations on balanced phylogenies also mirrored those
found on pure-birth trees. For \(\hat{\lambda}\), the mean value
increased with increasing input signal, but was negatively biased, and
was less than the input value across most of its range (Fig. S5: black
line). Additionally, the precision of \(\hat{\lambda}\) varied with
differing input levels, with the greatest variation found at
intermediate values of \(\lambda\) (Fig. S5 red line). Finally, the
distribution of \(\hat{\lambda}\) was not normal, and became more skewed
at more extreme values of \(\lambda\) (Fig. S5 blue line). \hfill\break

For \emph{K}, mean values increased with increasing phylogenetic signal,
though as was found with pure-birth trees, the increase was nonlinear
(Fig. S6 black line). Likewise, variation increased with increasing
phylogenetic signal (Fig. S6 red line), though the distribution of
\emph{K} was more normally distributed throughout its range, and across
different tree sizes, as compared with \(\hat{\lambda}\), though there
was some slight skew for high input levels of phylogenetic signal on
large phylogenies (Fig. S6 blue line). \hfill\break

With respect to effect sizes, both \(Z_{\lambda}\) and \(Z_K\) increased
with increasing input phylogenetic signal, but \(\hat{\lambda}\) was
strongly affected by tree size (fig.~S7). Also, \(Z_K\) increased more
linearly with increasing levels of phylogenetic signal, though there was
a slight effect of tree size with balanced phylogenies. But as before,
its standard deviation across input signal was more even across tree
sizes, implying more consistent precision.

@@

We also investigaged the effect of unresolved phylogenies on estimates
of phylogenetic signal by adjusting pure-birth trees to have 50\%
collapsed nodes, following the procedures of (1). Results from
simulations on phylogenies containing polytomies mirrored those found on
fully resolved trees. For \(\hat{\lambda}\), the mean value increased
with increasing input signal, but was negatively biased, and was less
than the input value across most of its range (Fig. S8: black line).
Additionally, the precision of \(\hat{\lambda}\) varied with differing
input levels, with the greatest variation found at intermediate values
of \(\lambda\) (Fig. S8 red line). Finally, the distribution of
\(\hat{\lambda}\) was not normal, and became more skewed at more extreme
values of \(\lambda\) (Fig. S8 blue line). \hfill\break

For \emph{K}, mean values increased with increasing phylogenetic signal,
though as was found with pure-birth trees, the increase was nonlinear
(Fig. S9 black line). Likewise, variation increased with increasing
phylogenetic signal (Fig. S9 red line), though the distribution of
\emph{K} was more normally distributed throughout its range, and across
different tree sizes, as compared with \(\hat{\lambda}\) (Fig. S9 blue
line). \hfill\break

With respect to effect sizes, both \(Z_{\lambda}\) and \(Z_K\) increased
with increasing input phylogenetic signal, but \(\hat{\lambda}\) was
strongly affected by tree size, whereas \(Z_K\) was more consistent
(Fig. S10). Also, \(Z_K\) increased more linearly with increasing levels
of phylogenetic signal, and its standard deviation across input signal
was more even across tree sizes, implying more consistent precision.
Thus, polytomies do not exert an appreciable effect on \(Z_K\).

@@

We also investigated the effect of incorporating \(\lambda\) in PGLS
analyses (phylogenetic regression and ANOVA). Here, patterns for
\(\hat\lambda\), \emph{K}, \(Z_\lambda\) and \(Z_\kappa\) were virtually
identical to those found previously. In essence, \(\hat\lambda\) was not
suitable as a test statistic representing phylogenetic signal due to all
properties previously shown (Figs. S11 \& S12), \emph{K} was more
appropriate, but associated non-linearly with signal strength (Figs. S13
\& 14), while \(Z_\kappa\) was found to be a superior effect size as
compared with \(Z_\lambda\) (Figs S15 \& S16). \hfill\break

With respect to model performance, input parameters (\(\beta\)) were
well estimated in phylogenetic regression and ANOVA when \(\lambda\) was
incorporated (Figs S17 \& S18), implying that \(\lambda\) may be used as
a tuning parameter for these models. Additionally, type I error and
statistical power were unaffected by the inclusion of \(\lambda\) (Figs
S19 \& S20), though power was reduced at smaller sample sizes in
phylogenetic ANOVA, mirroring prior findings of (2). These results
generally confirmed earlier findings of (3) for phylogenetic regression,
and extend them to the case of phylogenetic ANOVA.

@@

\textbf{Figure S11}. Response of Pagel's \(\lambda\) to increasing
strength of Brownian motion in phylogenetic regression (incorporating
\(\lambda\)). Gray line signifies the 1:1 line where the input value
matches the estimate.

@@

\textbf{Figure S12}. Response of Pagel's \(\lambda\) to increasing
strength of Brownian motion in phylogenetic ANOVA (incorporating
\(\lambda\)). Gray line signifies the 1:1 line where the input value
matches the estimate.

@@

\textbf{Figure S13}. Response of Blomberg's \textit{K} to increasing
strength of Brownian motion in phylogenetic regression (incorporating
\(\lambda\)). Gray line signifies the 1:1 line where the input value
matches the estimate.

@@

\textbf{Figure S14}. Response of Blomberg's \textit{K} to increasing
strength of Brownian motion in phylogenetic ANOVA (incorporating
\(\lambda\)). Gray line signifies the 1:1 line where the input value
matches the estimate.

@@

\textbf{Figure S15}. Response of effect sizes \(Z_{\lambda}\) and
\(Z_K\) to increasing strength of Brownian motion in phylogenetic
regression (incorporating \(\lambda\)).

@@

\textbf{Figure S16}. Response of effect sizes \(Z_{\lambda}\) and
\(Z_K\) to increasing strength of Brownian motion in phylogenetic ANOVA
(incorporating \(\lambda\)).

@@

\textbf{Figure S17}. Parameter estimates from phylogenetic regression
(incorporating \(\lambda\)). Gray line is the 1:1 line and red line is
the estimate.

\includegraphics[width=0.95\linewidth]{fig.S18}

\textbf{Figure S18}. Parameter estimates from phylogenetic ANOVA
(incorporating \(\lambda\)). Gray line is the 1:1 line and red line is
the estimate.

\includegraphics[width=0.95\linewidth]{fig.S19}

\textbf{Figure S19}. Power curves for phylogenetic regression
(incorporating \(\lambda\)). Type I error is found on the far-left of
the plot.

\includegraphics[width=0.95\linewidth]{fig.S20}

\textbf{Figure S20}. Power curves for phylogenetic ANOVA (incorporating
\(\lambda\)). Type I error is found on the far-left of the plot.

@@

\leavevmode\hypertarget{ref-AdamsCollyer2018b}{}%
2. Adams DC, Collyer ML (2018) Phylogenetic anova: Group-clade
aggregation, biological challenges, and a refined permutation procedure.
\emph{Evolution} 72(6):1204--1215.

\leavevmode\hypertarget{ref-Revell2010}{}%
3. Revell LJ (2010) Phylogenetic signal and linear regression on species
data. \emph{Methods in Ecology and Evolution} 1:319--329.
